---
title: "vincent.ines"
author: "Inès Vincent"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
date: "`r Sys.Date()`"
---
# Association between handgrip strength and functional performance in Interstitial Lung Disease patients

## 0. Project accessibility The project is available on GitHub at this link:
https://github.com/inesvincent/Project_R_Python_M2_REHAB.git

## 1. Scientific problem
### 1.1. Context

Interstitial lung diseases (ILD) are a heterogeneous group of pulmonary disorders characterized by diffuse involvement of the lung interstitium, leading to impaired gas exchange and symptoms such as dyspnea and cough. These conditions are associated with reduced exercise tolerance and decreased physical capacity, impacting daily life (Collège des Enseignants de Pneumologie, 2023) 

Although ILD primarily affects the respiratory system, it may indirectly alter overall physical function through reduced physical activity and deconditioning.

Handgrip strength is commonly used as a simple measure of upper-limb muscle strength and has been associated with functional performance and health outcomes (Bohannon, 2019; Norman et al., 2011). However, its ability to reflect overall muscle strength remains debated, with inconsistent associations reported depending on populations and methodologies (Szaflik et al., 2024).

Functional performance can be assessed using clinical tests such as walking speed and the five-times sit-to-stand (5STS) test, which reflect mobility, lower-limb strength, and overall functional capacity (Jones et al., 1999).

**What we are looking for:**  
To determine whether handgrip strength is associated with functional performance in patients with ILD, specifically with walking speed (4-meter walk test) and performance on the five-times sit-to-stand (5STS) test.


### 1.2. Aim
The aim of this study is to investigate the association between handgrip strength and functional performance in patients with ILD aged over 65 years, using walking speed and the five-times sit-to-stand (5STS) test as clinical measures

### 1.3. Method
This study is based on a monocentric database including 48 patients diagnosed with ILD. Data were collected during routine clinical assessments performed at the time of pulmonology consultation.

Handgrip strength was used to assess upper-limb muscle strength. Functional performance was evaluated using the 4-meter walk test and the five-times sit-to-stand (5STS) test. Walking speed was derived from the 4-meter walk time using the following formula:

Walking speed (m/s) = 4 / walk_time_4m

### 1.4. Participants
The study included 48 patients aged over 65 years diagnosed with interstitial lung disease (ILD), assessed during routine pulmonology consultations.

### 1.5. Outcome measure
The main variables analyzed were handgrip strength, walking speed derived from the 4-meter walk test, and performance on the five-times sit-to-stand (5STS) test.

---

## 2. Aim of the code
The objective of this code is to analyze the correlation between handgrip strength and functional performance in patients with ILD. The analysis includes the assessment of normality using graphical methods and the application of correlation tests (Pearson or Spearman, depending on data distribution) to evaluate the relationships between handgrip strength, walking speed, and performance on the five-times sit-to-stand (5STS) test.

---

## 3. Data organization

**File:** `frapid_data_clean.csv`  

**Columns:**  
- `patient_id` : Patient identifier  
- `age` : Age of the patient   
- `gender` : Sex of the patient  
- `handgrip_strength` : handgrip score :gives the muscular strength of the upper limbs (kg)  
- `walk_time_4m` : Time (in seconds) taken by the patient to walk 4 meters
- `time_5_raise_chair` : Time (in seconds) taken by the patient to perform 5 chair stands (5STS)
- `walking_speed_mps` : Walking speed (m/s), derived from the 4-meter walk test. 
$$
\text{Walking speed (m/s)} = \frac{4}{\text{walk time for 4m}}
$$
---

## 4. Statistical analysis

The data were first imported after being cleaned. The normality of the variables was assessed using QQ plots and verified with a Shapiro-Wilk test.
Based on this distribution, the choice of correlation test was adjusted: Pearson for normally distributed variables, and Spearman otherwise, as it is a nonparametric rank-based test.
These tests allow us to assess the existence of an association between the variables, with estimation of p-values.
Finally, a model adjusted for age and sex was developed to control for potential confounding factors.

### 4.1. Data import and preparation (section 5.1)
- **Aim:** To import the cleaned dataset and prepare variables for analysis.  
- **Input:** `frapid_data_clean.csv` in `data/clean` folder  
- **Output:** Dataset ready for statistical analysis, including derived variables such as walking speed.  

---

### 4.2. Assessment of normality (section 5.2)
- **Aim:** To assess the distribution of the variables using graphical methods (QQ-plots).  
- **Input:** Clean dataset  
- **Output:** Graphical evaluation of normality for handgrip strength, walking speed, and 5STS performance.  
---

### 4.3. Correlation analysis (section 5.3)
- **Aim:** To evaluate the correlation between handgrip strength and functional performance.
- **Input:** Clean dataset  
- **Calculation:** Correlation tests (Pearson or Spearman) depending on data distribution.  
- **Output:** Correlation coefficients and p-values indicating the strength and significance of the associations between:  
  - handgrip strength and walking speed  
  - handgrip strength and 5STS performance  

---

### 4.4. Adjusted analysis (section 5.4)  
- **Aim:** To adjust the associations for potential confounding variables (age and gender).  
- **Input:** Clean dataset  
- **Calculation:** Linear regression models including handgrip strength, age, and gender.  
- **Output:** Regression coefficients and statistical significance of adjusted associations.  

--- 

## 5. Code

### 5.0. Setup

This analysis was performed using R's basic functions. 
```{r}
#Only one additional package is required; if you haven't already, remove the comment symbol (#).
#install.packages("ggplot2")
```
```{r, message=FALSE, warning=FALSE}
#import necessary library
library(ggplot2)
```

### 5.1. Data import and preparation

The dataset was imported and checked to ensure correct structure and variable types.
```{r, message=FALSE, warning=FALSE}
# Import dataset
data <- read.csv("data/clean/frapid_data_clean.csv")

# Check structure
str(data)
summary(data)
```

### 5.2. Assessment of normality
```{r, message=FALSE, warning=FALSE}
# QQ-plot for normality
qqnorm(data$handgrip_strength, main = "QQ-plot: Handgrip strength")
qqline(data$handgrip_strength)

qqnorm(data$walking_speed, main = "QQ-plot: Walking speed")
qqline(data$walking_speed)

qqnorm(data$time_5_raise_chair, main = "QQ-plot: 5STS")
qqline(data$time_5_raise_chair)

```
A QQ-plot allows you to compare the distribution of observed data to a theoretical normal distribution: if the data points lie on the line, the data is normally distributed; conversely, deviations from this line indicate a deviation from normality.

According to the QQ plots obtained here, the distribution of the variables deviates from a normal distribution (particularly for the 5STS), which suggests the use of nonparametric tests. However, a Shapiro-Wilk test is performed to statistically confirm this hypothesis.

```{r, message=FALSE, warning=FALSE}
# Shapiro-Wilk test
shapiro.test(data$handgrip_strength)
shapiro.test(data$walking_speed)
shapiro.test(data$time_5_raise_chair)
```
The Shapiro-Wilk test is used to determine whether a variable follows a normal distribution. When the p-value > 0.05, the null hypothesis of normality is not rejected: the data are therefore considered to follow a normal distribution. Conversely, a p-value < 0.05 indicates that the data do not follow a normal distribution.

According to our results, grip strength (p = 0.8255) and walking speed (p = 0.1582) do not deviate significantly from a normal distribution. In contrast, the “sit-to-stand” test (p < 0.001) presents a non-normal distribution.

These results are consistent with the visual analysis of the QQ plots, where the data points for the 5STS deviate from the theoretical line, suggesting a non-normal distribution. Thus, although some variables are normally distributed, the presence of at least one non-normal variable calls for the use of nonparametric tests, such as Spearman’s correlation, to ensure the robustness and consistency of the analyses.


### 5.3. Correlation analysis

Spearman’s correlation was used to assess the association between handgrip strength and functional performance variables.
```{r, message=FALSE, warning=FALSE}
# Handgrip vs walking speed
cor.test(data$handgrip_strength, data$walking_speed, method = "spearman")

results_ws <- cor.test(data$handgrip_strength, data$walking_speed_mps, method = "spearman")

ggplot(data, aes(x = handgrip_strength, y = walking_speed_mps)) +
  geom_point(color = "darkgreen", size = 2, alpha = 0.7) +
  annotate("text",
           x = max(data$handgrip_strength, na.rm = TRUE),
           y = min(data$walking_speed_mps, na.rm = TRUE),
           label = paste0("Spearman rho = ", round(results_ws$estimate, 2),
                          "\np = ", signif(results_ws$p.value, 2)),
           hjust = 1, vjust = 0) +
  labs(title = "Handgrip vs Walking speed",
       x = "Handgrip strength (kg)",
       y = "Walking speed (m/s)") +
  theme_minimal()
```
A Spearman correlation analysis was performed to assess the association between grip strength and walking speed. A moderate, statistically significant positive correlation was observed (ρ = 0.43; p = 0.003), indicating that higher grip strength is associated with higher walking speed.

```{r, message=FALSE, warning=FALSE}
# Handgrip vs 5STS
cor.test(data$handgrip_strength, data$time_5_raise_chair, method = "spearman")

results_5STS <- cor.test(data$handgrip_strength, data$time_5_raise_chair, method = "spearman")

ggplot(data, aes(x = handgrip_strength, y = time_5_raise_chair)) +
  geom_point(color = "lightcoral", size = 2, alpha = 0.7) +
  annotate("text",
         x = max(data$handgrip_strength, na.rm = TRUE) * 0.95,
         y = max(data$time_5_raise_chair, na.rm = TRUE) * 0.98,
           label = paste0("Spearman rho = ", round(results_5STS$estimate, 2),
                          "\np = ", signif(results_5STS$p.value, 2)),
           hjust = 1, vjust = 1) +
  labs(title = "Handgrip vs 5STS",
       x = "Handgrip strength (kg)",
       y = "time_5_raise_chair (s)") +
  theme_minimal()
```
A Spearman correlation analysis was also performed to assess the association between grip strength and time on the 5STS test. A moderate, statistically significant negative correlation was observed (rho = -0.32; p = 0.029).

### 5.3.1 Interpretation of results

A Spearman correlation analysis was performed to assess the association between grip strength and functional performance. A moderate and significant positive correlation was found between grip strength and walking speed (ρ = 0.43; p = 0.003), indicating that higher grip strength is associated with higher walking speed.

Conversely, a moderate and significant negative correlation was observed between grip strength and time on the 5STS test (rho = -0.32; p = 0.029), suggesting that higher grip strength is associated with shorter completion times, which indicates better performance.

These results are based on unadjusted analyses; therefore, a supplementary analysis—adjusted for potential confounding factors such as age and sex—was conducted to assess the robustness of the observed associations.

### 5.4. Adjusted analysis
Linear regression analyses were conducted to assess the association between grip strength and functional performance, after adjustment for age and sex. The results of the models are presented below, along with 95% confidence intervals for the estimated coefficients.
```{r}
# Convert gender to factor for regression analysis
data$gender <- as.factor(data$gender)
```

```{r, message=FALSE, warning=FALSE}
model_walk <- lm(walking_speed_mps ~ handgrip_strength + age + gender, data = data)
summary(model_walk)

model_5sts <- lm(time_5_raise_chair ~ handgrip_strength + age + gender, data = data)
summary(model_5sts)

confint(model_walk)
confint(model_5sts)
```
### 5.4.1 Interpretation of results

Linear regression analyses were conducted to assess the association between grip strength and functional performance, taking age and sex into account. The β coefficient indicates the change in the outcome for a one-unit increase in the variable, holding all other factors constant. The 95% confidence intervals (95% CI) allow for an assessment of the precision of the estimates; an association is considered statistically significant when the interval does not contain 0 and the p-value is less than 0.05.

After adjustment, a significant positive association is observed between grip strength and walking speed (β = 0.02; 95% CI [0.001; 0.04]; p = 0.041). Age is also associated (β = -0.03; 95% CI [-0.052; -0.004]; p = 0.024), indicating that increasing age is linked to a decrease in walking speed. In contrast, no significant association is observed for sex (β = -0.07; 95% CI [-0.49; 0.34]; p = 0.72).

Furthermore, grip strength was no longer significantly associated with the 5STS test after adjustment (β = -0.11; 95% CI [-0.32; 0.09]; p = 0.28). Age remained significantly associated (β = 0.36; 95% CI [0.11; 0.62]; p = 0.006), indicating that older age is associated with a longer completion time. Gender is not significantly associated (β = -2.21; 95% CI [-6.56; 2.14]; p = 0.31).

## 6. Conclusion

This analysis aimed to assess the association between handgrip strength and functional performance in patients with ILD.

The analyses conducted reveal an association between handgrip strength and functional performance. A moderate positive correlation was observed with walking speed, as well as a negative correlation with the 5-step test, suggesting that greater muscle strength is associated with better performance.

After adjusting for age and sex, the association between handgrip strength and walking speed persists, indicating an independent effect of muscle strength. In contrast, the association with the 5STS test disappears, suggesting that it was partly influenced by age.

Furthermore, age appears to be a determining factor in functional performance, being associated with a decrease in walking speed and an increase in time on the 5STS, while sex does not appear to significantly influence these performance measures.
